Relationships in the extremes and their influence on competition and coexistence

### 2022-02-16 paper/Paper.Rnw
reviewing edits
comments for figure adjustment
fig comments
some edits to results

%***DAN: Methods paras here for G4
To achieve goal G4 from the Introduction, i.e., to provide an empirical example of species 
coexistence which highlights the role ATAs can play, we use a laboratory chemostat
system of freshwater diatoms explored by \cite{Descamps_2005}.
Those authors used measurements of growth rates of \emph{Cyclotella pseudostelligera} 
and \emph{Fragilaria crotonensis} over a range of temperatures to parametrize a resource
competition model. The model is 
\begin{align}
\frac{dS}{dt} &= D(S_0 - S) - Q_1x_1 \frac{V_1 S}{K_1 + S} - Q_2 x_2 \frac{V_2 S}{K_2 + S} \\
\frac{dx_j}{dt} &= x_j \frac{V_j S}{K_j+S}-D x_j, 
\end{align}
for $j=1,2$, where $S$ is a extracellular silicate concentration in the chemostat, $x_1$ is the population density
of \emph{F. crotonensis} and $x_2$ is that of \emph{C. pseudostelligera}, 
$D$ is the chemostat outflow rate, $S_0$ is concentration of silicates in the chemostat
inflow, the $Q_j$ are amounts of silicate per cell, the $V_j$ are maximum population growth rates, and
the $K_j$ are half-saturation constants for the dependence of growth rates on nutrient concentration.
The temperature-dependent parameters $Q_j$, $V_j$, and $K_j$ were obtained by 
\cite{Descamps_2005} through batch experiments and curve fitting or interpolation.
\cite{Descamps_2005} showed that permanent coexistence of the two species occurs when 
temperature fluctuates periodically, 
$\theta(t)=\theta_0 +  a\sin(2\pi t/P)$, 
with parameters $\theta_0 = 18^\circ$C, $a = 6^\circ$C, $P = 60$d. 
\cite{Ellner_2016, Ellner_2019} broke down coexistence into contributing mechanisms
according to (\ref{eq:previous_decomp}) for these same values of
$\theta_0$, $a$ and $P$. We further decomponse
coexistence according to (\ref{eq:full_decomp}), and we do so for ranges of values
of $\theta_0$, $a$ and $P$. In contrast to the lottery model case, for which simplifying assumptions
made it sufficient to decompose only the invasion growth rate of species 1, for this 
example we considered $\olr_{i \bs i} -q_{ij}\olr_{j \bs i}$ for both $i=1$, $j=2$ and for
$i=2$, $j=1$. Further details of the model setup are in SI section \ref{SI-sect:diatom_model}
and specifics of how our decomposition extends to this empirical case are in 
SI section \ref{SI-sect:Empircal_app}.
%***DAN: Jasmin, you had already created and started on the second of these SI sections, but the first one
%is new, and needs to be filled in. Please make a start at that when you get a chance. I've put some 
%notes there indicating what the content of that section should be.

%***DAN: Jasmin, I edited the below paragraph pretty heavily, please check,
%especially the quantitative statements at the end.
To display results, we plotted the terms of (\ref{eq:full_decomp}) against one of the parameters 
$\theta_0$, $a$ and $P$ at a time, while keeping the other two fixed at the values
used by \cite{Descamps_2005}. We also generated two-dimensional plots, where
two of the parameters $\theta_0$, $a$ and $P$ were varied and the third was kept fixed
at the value used by \cite{Descamps_2005}. For the two-dimensional plots, the quantity  
$\Delta_i^{[EC]}/\text{IGR}$ was displayed using color, to convey the 
importance of the ATA contribution relative to the whole IGR. Larger magnitudes
of this quantity indicate greater importance of ATAs. The case
$\Delta_i^{[EC]}>\text{IGR}>0$ (which produces $\Delta_i^{[EC]}/\text{IGR}>1$)
corresponds to ATA rescue, and the case $\Delta_i^{[EC]}<\text{IGR}<0$
(which also produces $\Delta_i^{[EC]}/\text{IGR}>1$) corresponds to ATA exclusion.
The condition $\Delta_i^{[EC]}/\text{IGR}>1$ encompasses the cases for which ATA
exclusion or rescue occurred, though the condition does not delineate, by itself,
which of these was the case.

@@

%G4
Fulfilling goal G4 of the Introduction (to provide an empirical example of species coexistence which highlights the role ATAs can play), Fig. \ref{fig:fig5} and Fig. \ref{fig:fig6} 
show that the ATA contributions have magnitude comparable to the 
magnitudes of other terms of the extended MCT decomposition and 
parameters subspaces exist where ATA rescue occurs. 
Within the investigated parameters of Fig.\ref{fig:fig5}, ATA contribution, $\Delta_i^{[EC]}$, 
is always positive, even in cases where IGR is negative, and is almost always second to 
the contribution of varying environment, $\Delta_i^{E}$, 
including for the original coexistence parameters specified in Gonzalez Decamps.  
Fig.\ref{fig:fig5} plots the various contributions of coexistence (see eq. \ref{eq:full_decomp}) for different varying parameters with two of the three parameters set constant in each plot. Increasing amplitude and mean temperature showed increasing trends in the IGR and decomposed quantities, while increasing periods showed less change.

@@

\begin{figure}
\includegraphics[width=.25\textwidth]{../results_figs/fig1_vertical.pdf}
\caption{Simple example of bivariate random variables with standard normal marginals showing
alternative patterns of association in the tails. The varables in (a) are ``left-tail 
associated,'' those in (b) are symmetrically associated, and those in (c) are ``right-tail associated''. All three cases have 
the same Pearson correlation, $\rho$, up to sampling variation, so Pearson correlation does not identify these very different 
patterns of association. We used asymmetrically tail associated cases (a, c) which are perfectly 
correlated in the lower or upper halves of the distributions, though it is also 
possible to generate asymmetrically tail associated noise with imperfect correlations in 
both halves of the distributions \citep{Ghosh_2020_AER,Nelsen2006}. The variables $(b_1,b_2)$ described in
the Introduction and Methods were obtained by generating 
noise as pictured and then multipling by $\sigma$ and adding $(\mu_1,\mu_2)$.\label{fig:noise}}
\end{figure}
%***DAN: Jasmin, I appologize for the following pedantic request, I explain below why it is necessary. Can you please change panel
%labels on this and all figure from the format "A", "B", "C" to the format "(a)", "(b)", "(c)" (i.e., lower case panel labels
%with parentheses around them). The reason for this is that this is Ecology Letters style. If we do not follow their style then their
%copy editors will make the change for us. I would prefer not to have copy editors messing around with our figures. Though Ecology Letters
%copy editors are usually pretty good, I have had many bad experiences with copy editors, generally, and I have had at least a few bad
%experiences with Ecology Letters copy editors. What happens is, they introduce errors into a paper while typesetting it, and one has to
%check their work very carefully to avoid those mistakes going to press. It's hard to blame them since it must be the most boring job 
%in the world, but anyway we do not want to have errors in our published paper as a result. The best way to prevent them introducing 
%errors into our figures is to not given them any reason to make any changes at all to our figures. The same change has to be 
%made to all main text and sup mat figures, and then all callouts to figures have to also be so changed, please. I will do these
%as much as possible while I edit, but if you could please also check it will help, and I am relying on you to change the panel labels
%in the figured themselves. I am sorry again for the tedium, here. Maybe some day they will improve the publication process so we do not 
%have to worry about nonsense like this.

@@

%***DAN: Jasmin, some changes are needed to the figure itself. Please make the text in the legend bigger.
%It looks as though a blue line is for ATA effects, and a red line is for the effects of correlation per se
%(that's what it says in the legend),
%but that differs from how you did it in the other figs. Or perhaps the legend is just wrong. Please check and
%fix so as to be consistent with the rest of the figures in which color represents which quantity and which
%notation represents which quantity. Also, please change panel labels to be of the format "(a)" instead of 
%"A". Make the panel labels bigger, as well as the axis numbers and labels. 
%***DAN: If I interpret correctly that the blue lines on Fig. 5 are the ATA contribution (needs to be 
%changed because you use a red line on other figs), there are large areas on panels a and c where the 
%IGR is between the ATA-contribution line and then x axis. Those are areas of ATA rescue (though you only
%know this after looking at the analogous plot for the other species). I think in 
%the next version of the figure you should consider highlighting them in a similar way to how you did 
%in Fig. 4. Those are a couple nice wide zones of ATA rescue! Let me know if you want to talk about this.
\begin{figure}
\includegraphics[width=\textwidth]{../results_figs/fig5.pdf}
\caption{The contribution of ATAs to species coexistence in the diatom system of \cite{Descamps_2005} when 
\emph{F. crotonensis} is the invader is meaningfully large compared to 
other components of the MCT decomposition. All panels show the IGR and the decomposition of
the IGR provided by the extended MCT, equation (\ref{eq:full_decomp}). 
%***DAN: This is for one species invading the other, right? Need to specify that, and say which is 
%invading which, and, at the end of the 
%caption indicate that the reverse invasion scenario is in sup mat and cite the corresponding figure there.
%***JASMIN: Done - ish. Need to make sup mat section 
For each panel, 
two of the three parameters defining the temperature oscillation (i.e., the amplitude, $a$; period, $P$; 
and mean, $\theta_0$; see Methods) are fixed at the original values used by 
\cite{Descamps_2005} ($a=6^\circ$C, $P=60$d, $\theta_0=18^\circ$C) and 
%***DAN: I added units for a  and \theta_0 here, please check these are correct 
%***JASMIN: Done 
the IGR and components of the MCT decomposition
are then plotted against the other parameter. The original value of the varying parameter 
is indicated by the vertical dashed magenta lines. ATA contributions 
had magnitude comparable to other components, 
%***DAN: Make sure this is still true after the correct version of C is put in, and after we
%resolve the potential problem with the wiggles on the plot. Reword if not.
indicating that ATAs can contribute meaningfully
to diatom coexistence in this system. In some instances, the IGR line (orange) falls 
between the ATA line (red)
and the horizontal axis, indicating that, for those parameter values, ATAs make the difference between species
coexistence and exclusion of the invader, \emph{F. crotonensis}. 
%***DAN: Not the weaker competitor, actually, right, but rather whichever competitor was being considered as the invader
%for these plots? If I'm right about that then please reword accordingly. 
%***JASMIN: Right, Done. 
Standard errors of plotted quantities ?.
%***DAN: Don't forget to fill in the standard error stuff
%***DAN: Here is where you add a statement citing another figure in the sup mat that takes the other
%species as the invader.
%***JASMIN: Done
Coexistence in this system and within 
our defined parameters only depends on the IGR of \emph{F. crotonensis} so we only 
show this case; though, an analogous plot (SI Fig. ) was generated for the reverse scenario where \emph{C. pseudostelligera} is the invader and 
\emph{F. crotonensis} is the resident.}  \label{fig:fig5}
\end{figure}

>>>>

%***DAN: Methods paras here for G4
To achieve goal G4 from the Introduction, i.e., to provide an empirical example of species 
coexistence which highlights the role ATAs can play, we use a laboratory chemostat
system of freshwater diatoms explored by \cite{Descamps_2005}.
Those authors used measurements of growth rates of \emph{Cyclotella pseudostelligera} 
and \emph{Fragilaria crotonensis} over a range of temperatures to parametrize a resource
competition model. The model is 
\begin{align}
\frac{dS}{dt} &= D(S_0 - S) - Q_1x_1 \frac{V_1 S}{K_1 + S} - Q_2 x_2 \frac{V_2 S}{K_2 + S} \\
\frac{dx_j}{dt} &= x_j \frac{V_j S}{K_j+S}-D x_j, 
\end{align}
for $j=1,2$, where $S$ is a extracellular silicate concentration in the chemostat, $x_1$ is the population density
of \emph{F. crotonensis} and $x_2$ is that of \emph{C. pseudostelligera}, 
$D$ is the chemostat outflow rate, $S_0$ is concentration of silicates in the chemostat
inflow, the $Q_j$ are amounts of silicate per cell, the $V_j$ are maximum population growth rates, and
the $K_j$ are half-saturation constants for the dependence of growth rates on nutrient concentration.
The temperature-dependent parameters $Q_j$, $V_j$, and $K_j$ were obtained by 
\cite{Descamps_2005} through batch experiments and curve fitting or interpolation.
\cite{Descamps_2005} showed that permanent coexistence of the two species occurs when 
temperature fluctuates periodically, 
$\theta(t)=\theta_0 +  a\sin(2\pi t/P)$, 
with parameters $\theta_0 = 18^\circ$C, $a = 6^\circ$C, $P = 60$d. 
\cite{Ellner_2016, Ellner_2019} broke down coexistence into contributing mechanisms
according to (\ref{eq:previous_decomp}) for these same values of
$\theta_0$, $a$ and $P$. We further decomponse
coexistence according to (\ref{eq:full_decomp}), and we do so for ranges of values
of $\theta_0$, $a$ and $P$. In contrast to the lottery model case, for which simplifying assumptions
made it sufficient to decompose only the invasion growth rate of species 1, for this 
example we considered $\olr_{i \bs i} -q_{ij}\olr_{j \bs i}$ for both $i=1$, $j=2$ and for
$i=2$, $j=1$. Further details of the model setup are in SI section \ref{SI-sect:diatom_model}
and specifics of how our decomposition extends to this empirical case are in 
SI section \ref{SI-sect:Empircal_app}.
%***DAN: Jasmin, you had already created and started on the second of these SI sections, but the first one
%is new, and needs to be filled in. Please make a start at that when you get a chance. I've put some 
%notes there indicating what the content of that section should be.
%***JASMIN: Working on filling out diatom model set up section

%***DAN: Jasmin, I edited the below paragraph pretty heavily, please check,
%especially the quantitative statements at the end.
%***JASMIN: Good 
To display results, we plotted the terms of (\ref{eq:full_decomp}) against one of the parameters 
$\theta_0$, $a$ and $P$ at a time, while keeping the other two fixed at the values
used by \cite{Descamps_2005}. We also generated two-dimensional plots, where
two of the parameters $\theta_0$, $a$ and $P$ were varied and the third was kept fixed
at the value used by \cite{Descamps_2005}. For the two-dimensional plots, the quantity  
$\Delta_i^{[EC]}/\text{IGR}$ was displayed using color, to convey the 
importance of the ATA contribution relative to the whole IGR. Larger magnitudes
of this quantity indicate greater importance of ATAs. The case
$\Delta_i^{[EC]}>\text{IGR}>0$ (which produces $\Delta_i^{[EC]}/\text{IGR}>1$)
corresponds to ATA rescue, and the case $\Delta_i^{[EC]}<\text{IGR}<0$
(which also produces $\Delta_i^{[EC]}/\text{IGR}>1$) corresponds to ATA exclusion.
The condition $\Delta_i^{[EC]}/\text{IGR}>1$ encompasses the cases for which ATA
exclusion or rescue occurred, though the condition does not delineate, by itself,
which of these was the case.

@@

%G4
Fulfilling goal G4 of the Introduction (to provide an empirical example of species coexistence which highlights the role ATAs can play), Fig. \ref{fig:fig5} and Fig. \ref{fig:fig6} 
show that the ATA contributions have magnitudes comparable to the 
magnitudes of other terms of the extended MCT decomposition and 
parameters subspaces exist where ATA rescue occurs. 
Fig.\ref{fig:fig5} plots the importance of various coexistence mechanisms for the 
persistence of \emph{F. crotonensis} and they're sum, the IGR
(see eq. \ref{eq:full_decomp}), for different varying temperature parameters 
($a$, $P$, and $\theta_0$) with two of the three parameters set constant in each plot.
Increasing amplitude and mean temperature showed increasing trends in the IGR and 
decomposed coexistence mechanisms, while increasing periodicity showed less change. 
Within the investigated parameters of Fig.\ref{fig:fig5}, ATA contribution, $\Delta_i^{[EC]}$, 
is always positive, even in cases where IGR is negative, and is almost always second to 
the contribution of varying environment, $\Delta_i^{E}$, 
including for the temperature parameters which \emph{F. crotonensis} 
and \emph{C. pseudostelligera} coexisted \emph{in vitro} ? \cite{Descamps_2005}.  
ATA rescue is revealed in Fig.\ref{fig:fig5} within parameter subspaces where the absence of 
ATA contribution would lead to exclusion of \emph{F. crotonensis}.
IGR decomposition of \emph{C. psuedostelligera} across the same subsets of
temperature parameters revealed that its IGR is always positive and the ATA contribution is 
not meaningful (see SI section). %Mean EC shows a rescue effect against variance in E

@@

\begin{figure}
\includegraphics[width=.25\textwidth]{../results_figs/fig1_vertical.pdf}
\caption{Simple example of bivariate random variables with standard normal marginals showing
alternative patterns of association in the tails. The varables in (a) are ``left-tail 
associated,'' those in (b) are symmetrically associated, and those in (c) are ``right-tail associated''. All three cases have 
the same Pearson correlation, $\rho$, up to sampling variation, so Pearson correlation does not identify these very different 
patterns of association. We used asymmetrically tail associated cases (a, c) which are perfectly 
correlated in the lower or upper halves of the distributions, though it is also 
possible to generate asymmetrically tail associated noise with imperfect correlations in 
both halves of the distributions \citep{Ghosh_2020_AER,Nelsen2006}. The variables $(b_1,b_2)$ described in
the Introduction and Methods were obtained by generating 
noise as pictured and then multipling by $\sigma$ and adding $(\mu_1,\mu_2)$.\label{fig:noise}}
\end{figure}
%***DAN: Jasmin, I appologize for the following pedantic request, I explain below why it is necessary. Can you please change panel
%labels on this and all figure from the format "A", "B", "C" to the format "(a)", "(b)", "(c)" (i.e., lower case panel labels
%with parentheses around them). The reason for this is that this is Ecology Letters style. If we do not follow their style then their
%copy editors will make the change for us. I would prefer not to have copy editors messing around with our figures. Though Ecology Letters
%copy editors are usually pretty good, I have had many bad experiences with copy editors, generally, and I have had at least a few bad
%experiences with Ecology Letters copy editors. What happens is, they introduce errors into a paper while typesetting it, and one has to
%check their work very carefully to avoid those mistakes going to press. It's hard to blame them since it must be the most boring job 
%in the world, but anyway we do not want to have errors in our published paper as a result. The best way to prevent them introducing 
%errors into our figures is to not given them any reason to make any changes at all to our figures. The same change has to be 
%made to all main text and sup mat figures, and then all callouts to figures have to also be so changed, please. I will do these
%as much as possible while I edit, but if you could please also check it will help, and I am relying on you to change the panel labels
%in the figured themselves. I am sorry again for the tedium, here. Maybe some day they will improve the publication process so we do not 
%have to worry about nonsense like this.
%***JASMIN: changed labels of figures. Done.

@@

%***DAN: Jasmin, some changes are needed to the figure itself. Please make the text in the legend bigger.
%It looks as though a blue line is for ATA effects, and a red line is for the effects of correlation per se
%(that's what it says in the legend),
%but that differs from how you did it in the other figs. Or perhaps the legend is just wrong. Please check and
%fix so as to be consistent with the rest of the figures in which color represents which quantity and which
%notation represents which quantity. Also, please change panel labels to be of the format "(a)" instead of 
%"A". Make the panel labels bigger, as well as the axis numbers and labels.

%***DAN: If I interpret correctly that the blue lines on Fig. 5 are the ATA contribution (needs to be 
%changed because you use a red line on other figs), there are large areas on panels a and c where the 
%IGR is between the ATA-contribution line and then x axis. Those are areas of ATA rescue (though you only
%know this after looking at the analogous plot for the other species). I think in 
%the next version of the figure you should consider highlighting them in a similar way to how you did 
%in Fig. 4. Those are a couple nice wide zones of ATA rescue! Let me know if you want to talk about this.
%***JASMIN: Fixed the line colors. Will block in the rescue areas. 
\begin{figure}
\includegraphics[width=\textwidth]{../results_figs/fig5.pdf}
\caption{The contribution of ATAs to species coexistence in the diatom system of \cite{Descamps_2005} when 
\emph{F. crotonensis} is the invader is meaningfully large compared to 
other components of the MCT decomposition. All panels show the IGR and the decomposition of
the IGR provided by the extended MCT, equation (\ref{eq:full_decomp}). 
%***DAN: This is for one species invading the other, right? Need to specify that, and say which is 
%invading which, and, at the end of the 
%caption indicate that the reverse invasion scenario is in sup mat and cite the corresponding figure there.
%***JASMIN: Done - ish. Need to make sup mat section 
For each panel, 
two of the three parameters defining the temperature oscillation (i.e., the amplitude, $a$; period, $P$; 
and mean, $\theta_0$; see Methods) are fixed at the original values used by 
\cite{Descamps_2005} ($a=6^\circ$C, $P=60$d, $\theta_0=18^\circ$C) and 
%***DAN: I added units for a  and \theta_0 here, please check these are correct 
%***JASMIN: Done 
the IGR and components of the MCT decomposition
are then plotted against the other parameter. The original value of the varying parameter 
is indicated by the vertical dashed magenta lines. ATA contributions 
had magnitude comparable to other components, 
%***DAN: Make sure this is still true after the correct version of C is put in, and after we
%resolve the potential problem with the wiggles on the plot. Reword if not.
indicating that ATAs can contribute meaningfully
to diatom coexistence in this system. In some instances, the IGR line (orange) falls 
between the ATA line (red)
and the horizontal axis, indicating that, for those parameter values, ATAs make the difference between species
coexistence and exclusion of the invader, \emph{F. crotonensis}. 
%***DAN: Not the weaker competitor, actually, right, but rather whichever competitor was being considered as the invader
%for these plots? If I'm right about that then please reword accordingly. 
%***JASMIN: Right, Done. 
Standard errors of plotted quantities ?.
%***DAN: Don't forget to fill in the standard error stuff
%***DAN: Here is where you add a statement citing another figure in the sup mat that takes the other
%species as the invader.
%***JASMIN: Done
Coexistence in this system and within 
our defined parameters only depends on the IGR of \emph{F. crotonensis} so we only 
show this case; though, an analogous plot (SI Fig. ) was generated for the reverse scenario where \emph{C. pseudostelligera} is the invader and 
\emph{F. crotonensis} is the resident.}  \label{fig:fig5}
\end{figure}
